Investigating magnetic dynamo evolution with TESS field dwarfs

### 2017-10-05 ms.tex
focussing on FFI stars

\input{dfm_stylez}
\pagestyle{myheadings}
\markright{\textsf{\footnotesize %
    Investigating rotational evolution with TESS field dwarfs}}

@@

The bimodal period distribution among field stars discovered by
\citet{mcquillan2013}, is shown in Figure \ref{fig:davenport}.
Recently, Co-I Davenport discovered this period bimodality extends throughout
all masses in the Kepler rotation sample for nearby stars
\citep{davenport2017}.
This feature could either reflect a drop in the star formation rate around 600
Myr ago or could be explained by a previously unknown variation in the
spin-down evolution for low-mass stars.
The TESS FFI and CTL targets will provide the ideal dataset to test these two
scenarios explaining the appearance of a period bimodality.
If the bimodal period distribution reflects a discontinuous age distribution,
the feature should be local and may disappear at greater distances or along
different lines of sight.
However, if the bimodality is truly due to a transition point in the spin-down
evolution at young ages, there should be little to no variation in the feature
with galactic position.
We will match the TESS FFI and CTL targets to the upcoming data
release from the Gaia mission \citep{perryman2001} to map the rotation period
distribution as a function of galactocentric position.
For the all-sky analysis we will focus on F and G stars since the 27.5 day
baseline of \TESS\ fields will limit measureable rotation periods to less than
around 15 days.
With the April 2018 data release from Gaia (DR2), we estimate that we will be
able to study rotation periods for G dwarfs in the TESS FFIs out to $\sim$3
kpc.
To characterize the rotation period bimodality as a function of galactic
position, we propose to analyse the light curves of 10,000 FFI and two-minute
targets that overlap with the Tycho-Gaia Astrometric Solution (TGAS) catalog,
selecting FGKM dwarfs with a range of galactic positions.
The FFI targets will be essential in order to reach $\sim$3kpc distances.
% Of these, the stars lying in the CVZs will be our highest priority.
M dwarfs rotation periods in the CVZz will allow us to study the rotation
period bimodality at lower masses, since the bimodality appears at longer
rotation periods for lower mass stars.
Unlike \kepler\ and \ktwo\ light curves, \TESS\ provides the all-sky coverage
necessary to map the period bimodality across the galaxy.

@@

Although the classical spin-down law of \citet{skumanich1972}: Period
$\propto$ Age$^1/2$ holds for all open clusters with measured periods, it does
not appear to describe old field stars \citep{angus2015, van-saders2016}.
\citet{van-saders2016} found that including a transition to a weakened magnetic
braking regime in the gyrochronology models, at a critical Rossby number,
provides an improved fit to the data.
Further calibration is needed however; a lack of rotation periods and reliable
ages for old and low-mass stars leaves the rotational evolution of stars older
than the Sun relatively unexplored.
\TESS\ will provide thousands of new rotation periods for both FFI and
two-minute cadence stars, with the maximum measurable period strongly
depending on the length of time it is observed for.
Of particular interest are the potentially long rotation periods measurable
for stars that fall in the Continuous Viewing Zones (CVZs).
Many of these will have spectra from the \TESS --\HERMES\ survey, from which it
will be possible to calculate isochronal ages, enabling further calibration of
the gyrochronology relations.
We will also use the rotation periods of TESS field dwarfs over the whole sky
to test the gyrochronology relations using galactic kinematics.
Many stars observable by \tess\ will have proper motions, parallaxes,
positions and radial velocities published in the second \Gaia\ data release.
These parameters provide the information necessary to calculate galactocentric
positions and action angles of the stars, both of which are age indicators.
% PI Angus is currently using the \kepler\ rotation periods of stars in the
% first \Gaia\ data release to calibrate the relations between rotation period
% and vertical action dispersion, two tracers of age.
% Since this relation is likely vary with both galactic longitude and latitude,
% the rotation periods produced from \TESS\ light curves will enable a
% comprehensive, all-sky calibration of these relations.
Since one of the few ways to accurately age-date fully convective, late M
dwarfs is via kinematics, these new relations will help to infer ages for all
stars with M$<$ 0.35M$_\odot$, to which gyrochronology cannot be applied.
% In order to calibrate gyrochronology using rotation periods of FGK dwarfs with
% \tess--\hermes\ spectra we propose to prioritize rotation periods of stars
% in the CVZs with spectra, followed by those without.

In addition, we will use the rotation periods of comoving stars identified in
the first \Gaia\ data release \citep{oh2016} to quantify the accuracy and
precision of gyrochronology.
We have identified 38 stars in 19 pairs with \Gaia\ proper motions and radial
velocities from \RAVE\ that are observable by \TESS, using {\tt tvguide}.
Of these, 34 stars in 17 pairs are already in the CTL list.
Because the stars in these pairs are assumed to be recently disrupted
binaries, we expect them to have the same age and measuring their rotation
periods will therefore provide a test of gyrochronology.

% Finally, we hope to use the improved gyrochronology relations, calibrated
% using \TESS\ data to infer the dependence of planet frequency on age and
% position in the galaxy.
% \kepler\ revealed tantalizing hints that exoplanets on short orbital periods
% have larger radii and are more infrequent around young stars \citep{mann2017a,
% mann2017b, rizzuto2017}.
% Do these trends continue into older ages and do they exist in the field?
% We will answer these questions by measuring rotation periods for all \TESS\
% field stars.

@@

The full set of available \TESS\ FFI and two-minute cadence light curves will
be exceedingly large and we intend to prioritize certain groups of stars for
period analysis, calculating a simple (and computationally fast) Lomb-Scargle
periodogram for the low priority objects and running a full probabilistic
period analysis on the high priority targets.
High priority targets include: comoving pairs, stars in the CVZs (especially
those with spectra), stars in TGAS, stars at large distances, plus any
potential asteroseismic dwarfs and exoplanet hosts.
Based on the expected photometric precision of \tess\ \citep{sullivan2015}, we
expect to be able to measure rotation periods for FFI stars down to 16th
magnitude.
We note, however that not all stars will have measurable rotation periods,
even if they are bright enough --- not all stars show rotational variability
in their light curves either because they are pole-on or because they are
inactive.
% Of these, the stars that also lie in the CVZs as well as the identified \Gaia\
% comoving pairs will be our highest priority.
% Our next priority will be stars in the CVZs, with spectra, that do not appear
% in TGAS, followed by those without spectra.
% M dwarfs rotation periods in the CVZz will allow us to study the rotation
% period bimodality at lower masses, since the bimodality appears at longer
% rotation periods for lower mass stars.
% We will then focus on measuring rotation periods for a sample of stars that is
% representative of the full set of FFI and CTL targets in order to enable
% exoplanet population studies.
% An large amount of computational resources are available to us, with access to
% the Columbia supercomputer as well as facilities at the Flatiron Institute.
% We expect that each target will require around one minute of computation time
% and have access to around 10,000 computing hours.
% We therefore expect to analyze around 60,000 light curves.

@@

\paragraph{Expected Impact}
Initially we will provide rotation period posterior probability density
functions and best fit estimates for around 30,000 high priority two-minute
cadence and FFI targets.
We will then produce a catalog of rotation periods for lower priority stars
using a combination of Lomb-Scargle periodogram and autocorrelation
techniques.
Finally, we will release a catalog of rotation periods for two-minute cadence
targets.
We will also release an open source python package for measuring rotation
periods of both FFI and two-minute cadence targets.

>>>>

\input{dfm_stylez}
\pagestyle{myheadings}
\markright{\textsf{\footnotesize %
    Investigating rotational evolution with TESS FFI field dwarfs}}

@@

The bimodal period distribution among field stars discovered by
\citet{mcquillan2013}, is shown in Figure \ref{fig:davenport}.
Recently, Co-I Davenport discovered this period bimodality extends throughout
all masses in the Kepler rotation sample for nearby stars
\citep{davenport2017}.
This feature could either reflect a drop in the star formation rate around 600
Myr ago or could be explained by a previously unknown variation in the
spin-down evolution for low-mass stars.
The TESS FFI targets will provide the ideal dataset to test these two
scenarios explaining the appearance of a period bimodality.
If the bimodal period distribution reflects a discontinuous age distribution,
the feature should be local and may disappear at greater distances or along
different lines of sight.
However, if the bimodality is truly due to a transition point in the spin-down
evolution at young ages, there should be little to no variation in the feature
with galactic position.
We will match the TESS FFI targets to the upcoming data
release from the Gaia mission \citep{perryman2001} to map the rotation period
distribution as a function of galactocentric position.
For the all-sky analysis we will focus on F and G stars since the 27.5 day
baseline of \TESS\ fields will limit measureable rotation periods to less than
around 15 days.
With the April 2018 data release from Gaia (DR2), we estimate that we will be
able to study rotation periods for G dwarfs in the TESS FFIs out to $\sim$3
kpc.
To characterize the rotation period bimodality as a function of galactic
position, we propose to analyse the light curves of 10,000 FFI and two-minute
targets that overlap with the Tycho-Gaia Astrometric Solution (TGAS) catalog,
selecting FGKM dwarfs with a range of galactic positions.
The FFI targets will be essential in order to reach $\sim$3kpc distances.
% Of these, the stars lying in the CVZs will be our highest priority.
M dwarfs rotation periods in the CVZz will allow us to study the rotation
period bimodality at lower masses, since the bimodality appears at longer
rotation periods for lower mass stars.
Unlike \kepler\ and \ktwo\ light curves, \TESS\ provides the all-sky coverage
necessary to map the period bimodality across the galaxy.

@@

Although the classical spin-down law of \citet{skumanich1972}: Period
$\propto$ Age$^1/2$ holds for all open clusters with measured periods, it does
not appear to describe old field stars \citep{angus2015, van-saders2016}.
\citet{van-saders2016} found that including a transition to a weakened magnetic
braking regime in the gyrochronology models, at a critical Rossby number,
provides an improved fit to the data.
Further calibration is needed however; a lack of rotation periods and reliable
ages for old and low-mass stars leaves the rotational evolution of stars older
than the Sun relatively unexplored.
\TESS\ will provide thousands of new rotation periods for both FFI and
two-minute cadence stars, with the maximum measurable period strongly
depending on the length of time it is observed for.
Of particular interest are the potentially long rotation periods measurable
for stars that fall in the Continuous Viewing Zones (CVZs).
Many of these will have spectra from the \TESS --\HERMES\ survey, from which it
will be possible to calculate isochronal ages, enabling further calibration of
the gyrochronology relations.
We will also use the rotation periods of TESS field dwarfs over the whole sky
to test the gyrochronology relations using galactic kinematics.
Many stars observable by \tess\ will have proper motions, parallaxes,
positions and radial velocities published in the second \Gaia\ data release.
These parameters provide the information necessary to calculate galactocentric
positions and action angles of the stars, both of which are age indicators.
Since one of the few ways to accurately age-date fully convective, late M
dwarfs is via kinematics, these new relations will help to infer ages for all
stars with M$<$ 0.35M$_\odot$, to which gyrochronology cannot be applied.

In addition, we will use the rotation periods of comoving stars identified in
the first \Gaia\ data release \citep{oh2016} to quantify the accuracy and
precision of gyrochronology.
We have identified 38 stars in 19 pairs with \Gaia\ proper motions and radial
velocities from \RAVE\ that are observable by \TESS, using {\tt tvguide}.
Of these, 34 stars in 17 pairs are in the CTL list.
Since these stars would make interesting targets for both asteroseismology and
exoplanet search, we include a list of them with this proposal.
The stars in these pairs are assumed to be recently disrupted binaries so we
expect them to have the same age; measuring their rotation periods will
therefore provide a test of gyrochronology.

@@

The full set of available \TESS\ FFI light curves will be exceedingly large
and we intend to prioritize certain groups of stars for period analysis,
calculating a simple (and computationally fast) Lomb-Scargle periodogram for
the low priority objects and running a full probabilistic period analysis on
the high priority targets.
High priority targets include: comoving pairs, stars in the CVZs (especially
those with spectra), stars in TGAS, stars at large distances, plus any
potential asteroseismic dwarfs and exoplanet hosts.
Based on the expected photometric precision of \tess\ \citep{sullivan2015}, we
expect to be able to measure rotation periods for FFI stars down to 16th
magnitude.
We note, however that not all stars will have measurable rotation periods,
even if they are bright enough --- not all stars show rotational variability
in their light curves either because they are pole-on or because they are
inactive.

@@

\paragraph{Expected Impact}
Initially, we will provide rotation period posterior probability density
functions and best fit estimates for around 30,000 high priority FFI targets.
We will then produce a catalog of rotation periods for lower priority FFI
stars using a combination of Lomb-Scargle periodogram and autocorrelation
techniques.
Finally, we will release a catalog of rotation periods for two-minute cadence
targets.
The full catalog will likely contain several million rotation periods.
We will also release an open source python package for measuring rotation
periods of both FFI and two-minute cadence targets.
